## Functions and Conditionals

### Conditionals

The basic syntax for an if statement goes like this:

```{r}
if (logical statement) {
  do something
} else {
  do something else
}
```

The logical statements should boil down to a boolean; for example:

```{r}
x <- 10

if (x == 10) {
  print("X is equal to 10")
}
else {
  print("X is not equal to 10")
}
```

We can also add else if statements in the middle if we need more options.

```{r}
if (x <= 5) {
  print(x)
} else if (x == 7) {
  print("x is equal to 7")
} else {
  print("x is greater than 5 and not equal to 7")
}
```

Neat! We can also nest conditionals just like we do in Python.

```{r}
if(x == 10) {
  if(x != 2) {
    print("x is still equal to 10")
  }
}
```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}
for(i in 1:5) {
  print(i)
}
```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}
a_vector <- c(1,1,1,1,1,1,1)

for(i in c(1,3,6)) {
  a_vector[i] <- 6
}

a_vector
```

We can loop over more than just lists:

```{r}
mat <- matrix(1, nrow = 2, ncol = 2)

for(row in 1:2) {
  for(col in 1:2) {
    mat[col, row] <- 0
    print(mat)
  }
}
```

Like Python, we have while loops in R:

```{r}
b <- 0
while(b < 10) {
  b <- b + 2
  print(b)
}
```

## Functions

When should we use functions?

```{r}
our_function <- function(optional_params){
  # do some stuff
}
```

An example:

```{r}
test_temp <- c(16.55, 26.94, 32.59, 8.58, 28.00, 28.54, 20.97, 21.17, 21.04, 18.76)

c_to_f <- function(dat) {
  temp_fahr <- dat * (9/5) + 32
  return(temp_fahr)
}

print(c_to_f(26))
```

This is all cool.. but it didn't do anything. We need to call the function!

```{r}
data_fahr <- c_to_f(test_temp)
data_fahr
```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}
test_mat <- matrix(c(1,1,1,2,2,2), nrow <- 2)
apply(test_mat, 1, sum)
```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}
test_temp_mat <- matrix(c(16.55, 26.94, 32.59, 8.58))

apply(test_temp_mat, 1, c_to_f)
```

### Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    ```{r}
    # Cat years, Dog years:  Convert human years into cat years and dog years.
    human_years_cat_years_dog_years <- function(human_years){
      cat_years <- 0
      dog_years <- 0
      years <- human_years
      
      while(years > 2) {
        cat_years <- cat_years + 4
        dog_years <- dog_years + 5
        years <- years - 1
      }
      
      if(years == 2) {
        cat_years <- cat_years + 9
        dog_years <- dog_years + 9
        years <- years - 1
      }
      
      cat_years <- cat_years + 15
      dog_years <- dog_years + 15
      
      return(c(human_years, cat_years, dog_years))
    }


    # Quarter of the year: given a month, return what quarter of the year it resides in
    quarter_of <- function(month){
       if(month < 4) {
         return(1)
       } else if(month < 7) {
         return(2)
       } else if(month < 10) {
         return(3)
       } else {
         return(4)
       }
    }


    # Find maximum and minimum values of a list
    minimum <- function(vec){
      vec <- sort(vec)
      return(vec[1])
     } 

    maximum <- function(vec){
      vec <- sort(vec)
      return(vec[length(vec)])
    } 


    # Invert values: given a set of numbers, return he additive inverse of each
    invert <- function(vec){
      vec_2 <- c(-1)
      vec <- vec * vec_2
      return(vec)
    }


    # Convert a string to a number
    string_to_number <- function(x){
      return(as.numeric(x))
    }
    ```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}
    fb <- function(x) {
      if(x %% 3 == 0 & x %% 5 == 0) {
        return("fizzbuzz")
      }
      if(x %% 3 == 0) {
        return("fizz")
      }
      if(x %% 5 == 0) {
        return("buzz")
      }
    }

    my_seq <- 1:30
    for(i in my_seq) {
      print(fb(i))
    }
    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}
    # You may feel more comfortable figuring this out in Python first, then translating into R later.

    marioStairs <- function(num) {
      steps = ""

      for(i in 1:num) {
        steps = paste(steps, "X")
        print(steps)
      }
    }

    marioStairs(5)
    ```
